Bumble bee competition revisited under global climate change

### 2020-05-23 manuscript.Rmd
oh yeah

Bumble bees (Hymenoptera: *Bombus* spp.), due to their well-studied natural history and amenability to field observation, have become a model system for the study of biotic interactions in ecological communities. In particular, the broad and overlapping distributions of bumble bee species, together with their gross similarity in morphology, behavior, and dietary niche, have attracted the attention of researchers interested in questions of interspecific competition and, more broadly, the ecological constraints of species coexistence [@Plowright1984-lc; @Goulson2008-zy]. In this regard, however, bumble bees have proven somewhat enigmatic [@Ranta1981-vx]. On the one hand, there is evidence that resource limitation, a necessary precondition for competition, can be a significant constraint on local bumble populations [@Heinrich1976-wc], even to the point of widespread colony failure due to starvation [@Bowers1985-df]. Moreover, the species of certain bumble bee communties have been shown to be over-dispersed in terms of tongue length (a matching trait corresponding to floral corolla depth) or phyologeny (understood to encompass unmeasured trait variation), providing evidence for competition as a community assembly process [e.g. @Inouye1977-kl; @Pellissier2013-pk]. Nevertheless, it is equally evident that sympatric bumble bee species do not *always* compete for floral resources to the point of competitive exclusion, since bumble bee communities with an apparent excess of species richness and deficit of trait (or phylogenetic) dispersion are commonly observed [@Goulson2008-zy].

To account for this inconsistency, @Ranta1981-vx proposed that bumble bees communities are governed by a spatiotemporally heterogeneous "mosaic" of competitive conditions, arising from the interaction of bumble bee species' traits and nest locations with spatiotemporal variation in floral resource abundance and composition, such that competitive inequalities between species are sufficiently variable (and fine-grained relative to the scale of observation) to allow more species to coexist than predicted by trait-based competition models. A complementary explanation, based on the core/satellite species hypothesis [@Hanski1982-hs], is that competitive forces are significant only among regionally abundant and widespread "core species", and the effects of competition are more pronounced in the structure of regional species pools than in that of local assemblages [@Hanski1982-xd].

As pointed out by @Williams1989-vr, however, the question of competition as a process of community assembly, relevant to evolutionary and biogeographic scales of inquiry, differs from that of competition as a behavioral phenomenon within a given community, irrespective of the forces responsible for its assembly.

"This is not to deny that interspecific competition affects patterns of foraging by bumble bees, which has been demonstrated by the effect of competitor removal  (e.g. Inouye, 1978; Plowright, Pendrel & McLaren, 1978; Plowright & Rodd, 1980; Pleasants,  1981). It is  concluded only that interspecific competition does not appear to govern the composition of local  assemblages of bumble bee species in the manner predicted by  the  theory of species-packing." [@Williams1989-vr]

Bring up Hanksi's core/satellite hypothesis.

Steep elevation gradients can serve as space-for-time proxies in the study of climate effects, capturing a broad range of climate variation  while minimizing the confounding effects of geographic distance [e.g. @Hoiss2015-gv].

@@

When phylogenetic/trait dispersion are used as indices of competition, the relevant question is whether competition played a role in community assembly. These indices do not answer the question of whether, right now, competition is happening. Trait-matching, niche overlap, specialization (e.g. d') based indices are better suited for the question of immediate competiion. It could be that competition was not important over the time scales relevant to community assembly, but now it has become a significant phenomenon due to some change in the system. Or, it is possible that competition is spatiotemporally transient, and its net effect is not sufficient to cause a trait or phylogenetic signal.

Bumble bees (Hymenoptera: *Bombus* spp.) are classic model organisms for the study of resource competition in biological communities. Nevertheless, the role of resource competition in structuring bumble bee communities has proven difficult to generalize. Resource limitation, a precondition for competition,

@Heinrich1976-wc found that a bumble bee community can deplete the daily standing nectar pool of a given locality by more than 90%, and starvation has been implicated in the demise of colonies in isolated subalpine meadows [@Bowers1985-df], in. Other studies have found local bumble bee communities to be over-dispersed in terms of tongue length [e.g. @Inouye1977-kl], a matching trait [sensu @Schleuning2020-vq] corresponding to floral corolla depth, and phylogeny [@Pellissier2013-pk], which can be interpreted as a catch-all for unmeasured trait variation. Nevertheless, it is equally evident that sympatric bumble bee species do not *always* compete for floral resources to the point of competitive exclusion, since bumble bee communities with an apparent excess of species richness and deficit of trait dispersion are commonly observed [@Goulson2008-zy].

>>>>

Bumble bees (Hymenoptera: *Bombus* spp.) have become a model system for the study of biotic interactions in ecological communities. In particular, the broad and overlapping distributions of bumble bee species, together with their gross similarity in morphology, behavior, and dietary niche, have attracted the attention of researchers interested in questions of interspecific competition and, more broadly, the rules of community assembly and species coexistence [@Plowright1984-lc; @Goulson2008-zy]. In this regard, however, bumble bees have proven somewhat enigmatic. On the one hand, there is evidence that resource limitation can be a significant constraint on local bumble populations [@Heinrich1976-wc; @Bowers1985-df], and certain bumble bee communties have been shown to be over-dispersed in terms of tongue length (a matching trait corresponding to floral corolla depth) or phyologeny (understood to encompass unmeasured trait variation), providing evidence for competition as a community assembly process [e.g. @Inouye1977-kl; @Pellissier2013-pk]. Nevertheless, bumble bee communities with an apparent excess of species richness and deficit of trait (or phylogenetic) dispersion are commonly observed [@Ranta1981-vx]. To account for this inconsistency, @Ranta1981-vx proposed that bumble bees communities are governed by a spatiotemporally heterogeneous "mosaic" of competitive conditions, such that competitive inequalities between species are sufficiently variable to allow more species to coexist than predicted by trait-based competition models. A complementary explanation, based on the core/satellite species hypothesis [@Hanski1982-hs], is that competitive forces are significant only among regionally abundant and widespread "core species", and the effects of competition are more pronounced in the structure of regional species pools than in that of local assemblages [@Hanski1982-xd].

As pointed out by @Williams1989-vr, however, the question of competition as a biogeographic/evolutionary-scale driver of communty assembly differs from that of competition as a local/behavioral-scale driver of species' interactions. While the role of competition in bumble bee community assembly remains somewhat obscure, there is broad empirical evidence that bumble bees facultatively partition resources on the basis of trait-matching and possibly dominance hierarchy [e.g @Inouye1978-qi; @Morse1982-bh; @Ishii2013-sm; @Brosi2013-mu]. This distinction between spatiotemporal scales of competition, echoing the logic of @Ranta1981-vx, becomes especially important when the scope of inquiry switches from retrospective to prospective, and all the more so when there is reason to think that the ecological context of competition is undergoing rapid and unprecedented change. Since the height of bumble bee competition research in the 1980s, such a shift has indeed occurred, with retrospective questions of community assembly eclipsed by the urgency of conservation in the face of global changes in climate and land use [@Goulson2010-lo]. Given the uncertain spatiotemporal lags and exogenous drivers of community assembly, inferring the past role of competition through the study of bumble bee community structure may be of limited use in predicting the future fitness and distribution of bumble bee species, together with the ecological functions and services they provide. Studying local/behavioral-scale competition in bumble bee communities along climate gradients provides a vital complement to the classical study of bumble bee community structure.

The study of competition at local/behavioral scales, however, entails two empirical challenges. First, it is not enough simply to know which species occur and their similarity in traits/phyologeny; one must record their behavior with respect to their shared resources, and at a frequency appropriate to the temporal scale of inquiry. Second, climate variation must be decoupled from confounding spatial effects. The use of species interaction networks along steep elevation gradients provides an elegant solution to both of these challenges.

Interaction networks provide a tractable analytical framework for the study of competition and resource partitioning [], and steep elevation gradients can serve as space-for-time proxies in the study of climate effects, capturing a broad range of climate variation while minimizing the confounding effects of geographic distance [e.g. @Hoiss2015-gv].

@@
